---
title: "MSE1_Drought"
output: html_document
date: "2026-05-26"
---
Drought treatment in MSE1 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages 
library(ggplot2)
library(readxl)
library(dplyr)
library(nlme)
library(lme4)
library(lmerTest)

# set theme 
set_theme(theme_bw())
```

```{r}
# load data 
x <- read_xlsx("./SoilProbe.xlsx")

x$Stress_ID <- as.factor(x$Stress_ID)
x$Drought <- as.factor(x$Drought)
```
In the experiment, we only included soil moisture probes in one pot per block (and note that these are not true "blocks". This was a completely random design and I call the racks that the pots were stacked in "blocks"). Probes were also placed so that each unique treatment had one probe measuring soil moisture. I collected soil moisture measurements every two days immediately post-watering. For treatments that were droughted, I collected soil probe readings after watering the appropriate treatments. 

```{r}
# plot soil moisture across all probes and the full experiment 
soilmoisture <- ggplot(x, aes(x = Time, y = Probe_Reading, color = Drought, group = ID)) + 
  geom_line() + 
  labs(x = "Days Since Planting", 
       y = "Soil Volumetric Water Content (%)", 
       color = "Drought Treatment") + 
  geom_vline(xintercept = 16, 
             linetype = "dashed") + 
    scale_color_manual(values = c("Presence" = "#e6194B", 
                                "Absence" = "#4363d8"))

ggsave("./soilmoisture.pdf", soilmoisture)
```
Note Timepoint 21 was between waterings, whereas all other measurements were immediately post-watering 

```{r}
# comparing drought vs control for just the drought portion of the experiment
soil_drought <- x %>%
  filter(Time %in% 16:21)

droughtvscontrol <- ggplot(soil_drought, aes(x = Drought, y = Probe_Reading, color = Drought)) + 
  geom_boxplot() + 
  geom_point(alpha = 0.5, size = 2) + 
  scale_color_manual(values = c("Presence" = "#e6194B", 
                                "Absence" = "#4363d8")) + 
  labs(x = "Treatment", 
       y = "Soil Volumetric Water Content (%)")

ggsave("./droughtvscontrol.pdf", droughtvscontrol)
```

```{r}
# stats
soil_drought
t.test(Probe_Reading ~ Drought, data = soil_drought)
```
