---
title: "MSE1_Univariate_Multiplicative_v2"
output: html_document
date: "2026-05-15"
---
---
title: "MSE1_Univariate_Multiplicative"
author: "Natalie Ford"
date: "`r Sys.Date()`"
output: html_document
---
Final R Markdown file for the biomass and root architecture multivariate analysis for MSE1 publication

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE)
```

```{r}
# Load packages & set theme 
library(dplyr); packageVersion("dplyr") #1.2.0 
library(effsize); packageVersion("effsize") # 0.8.1
library(ggplot2)
library(ggpattern)
library(readxl)
library(tidyr)
library(purrr)
library(stringr)
library(ggpubr)
library(forcats)

theme_set(theme_bw())
```

```{r}
# Load and prep data 
x <- read_xlsx("~/MSE1_metadata.xlsx")

# remove sample IDs 59 and 114 because they did not survive the experiment and there is no data associated with them 
y <- x %>% filter(ID !=59)
mse1 <- y %>% filter(ID !=114)

# remove sample IDs 71 & 66 (outliers in root architecture)
mse1 <- mse1 %>% filter(ID!=71)
mse1 <- mse1 %>% filter(ID!=66)

# remove sample IDs 47 & 164 (outliers in microbiome data)
mse1 <- mse1 %>% filter(ID!=47)
mse1 <- mse1 %>% filter(ID!=164)

# remove samples that have no exudate data 
mse1 <- mse1 %>% filter(ID!=67)
mse1 <- mse1 %>% filter(ID!=80)
mse1 <- mse1 %>% filter(ID!=125)
mse1 <- mse1 %>% filter(ID!=128)

# make sample ID, number of stressors, and stress ID factors 
mse1$ID <- as.factor(mse1$ID)
mse1$Num_Stress <- as.factor(mse1$Num_Stress)

#change the order of the factor levels explicitly for plots 
mse1$Stress_ID <- factor(mse1$Stress_ID, 
                            levels = c("Control", "MP", "Drought", "Herbicide", "Antibiotic",
                                       "Drought_MP", "Herbicide_MP", 
                                       "Antibiotic_MP", "Drought_Herbicide",
                                       "Antibiotic_Drought","Antibiotic_Herbicide", 
                                       "Drought_Herbicide_MP", "Antibiotic_Drought_MP", 
                                       "Antibiotic_Herbicide_MP", "Antibiotic_Drought_Herbicide", 
                                       "Antibiotic_Drought_Herbicide_MP"))
```

# Biomass
## Shoot Biomass ##
#### Calculating the multiplicative null model and effect sizes 
```{r}
analyze_all_combos <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_sbiomass = mean(Shoot_Biomass, na.rm = TRUE), 
      sd_sbiomass = sd(Shoot_Biomass, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_sbiomass
    Xc <-  control$mean_sbiomass
    X_obs <- combo_data$mean_sbiomass
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data$sd_sbiomass^2 / (combo_data$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_sbiomass^2 / (single_data$n * single_data$mean_sbiomass^2)) +  # Singles
    (k - 1) * (control$sd_sbiomass^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results <- map_dfr(combinations, ~calculate_effect_size(.x, treatment_summaries))
  return(results)
} 

# run it 
results <- analyze_all_combos(mse1)
print(results)
```

#### Plotting the observed effect sizes for each stress combination treatment 
```{r}
plot_data <- results %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

shoot_bio_multi_effects <- ggplot(plot_data, aes(x = reorder(combination, n_stressors), y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment", y = "Effect Size (95% CI)",
       title = "Shoot Biomass",
       color = "Interaction Type")
```

#### Plotting raw data 
```{r}
# remove NAs
mse1_clean <- mse1 %>% drop_na(Stress_ID)

# figure out what the mean of the control is to add the horizontal dotted line 
tapply(mse1_clean$Shoot_Biomass, mse1_clean$Stress_ID, summary)
#median shoot biomass for control = 179.3
#mean shoot biomass for control = 178.6

# plot
shoot_biomass_boxplot <- ggplot(mse1_clean, aes(x = Stress_ID, y = Shoot_Biomass, color = Stress_ID)) + 
  geom_boxplot(outlier.alpha = 0.4) + 
  geom_point(size = 2, alpha = 0.4) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept=179.3, linetype = "dotted") + 
    scale_color_manual(
    values = c(
        "Control" = "#e6194B",
        "MP" = "#3cb44b",
        "Drought" = "#ffe119",
        "Herbicide" = "#4363d8",
        "Antibiotic" = "#f58231",
        "Drought_MP" = "#42d4f4",
        "Antibiotic_MP" = "#f032e6",
        "Herbicide_MP" = "#469990",
        "Antibiotic_Herbicide" = "#9A6324",
        "Drought_Herbicide" = "#800000",
        "Antibiotic_Drought" = "#911eb4",
        "Antibiotic_Drought_Herbicide" = "#bfef45", 
        "Antibiotic_Drought_MP" = "#808000",
        "Antibiotic_Herbicide_MP" = "#000075",
        "Drought_Herbicide_MP" = "#a9a9a9", 
        "Antibiotic_Drought_Herbicide_MP" = "#000000"
    )) + 
  labs(y = "Shoot Biomass", x = "Treatment") + 
  theme(legend.position = "none")

ggsave("~/shoot_biomass_boxplot.pdf", shoot_biomass_boxplot)
```
## Root Biomass ##
#### Calculating the multiplicative null model estimates and effect sizes
```{r}
analyze_all_combos_rbiomass <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_rbiomass = mean(Root_Biomass, na.rm = TRUE), 
      sd_rbiomass = sd(Root_Biomass, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_rbiomass <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_rbiomass <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_rbiomass) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_rbiomass
    Xc <-  control$mean_rbiomass
    X_obs <- combo_data_rbiomass$mean_rbiomass
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_rbiomass$sd_rbiomass^2 / (combo_data_rbiomass$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_rbiomass^2 / (single_data$n * single_data$mean_rbiomass^2)) +  # Singles
    (k - 1) * (control$sd_rbiomass^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_rbiomass <- map_dfr(combinations, ~calculate_effect_size_rbiomass(.x, treatment_summaries))
  return(results_rbiomass)
} 

# run it 
results_rbiomass <- analyze_all_combos_rbiomass(mse1)
print(results_rbiomass)
```

#### Plotting effect sizes (multiplicative model)
```{r}
plot_data_rbiomass <- results_rbiomass %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_bio_multiplicative_effects <- ggplot(plot_data_rbiomass, aes(x = reorder(combination, n_stressors), y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment", y = "Effect Size (95% CI)",
       title = "Root Biomass",
       color = "Interaction")
```

#### Plotting raw data 
```{r}
# figure out what the mean of the control is to add the horizontal dotted line 
tapply(mse1_clean$Root_Biomass, mse1_clean$Stress_ID, summary)
#median root biomass for control = 71.40
#mean root biomass for control = 85.78

# plot
root_biomass_boxplot <- ggplot(mse1_clean, aes(x = Stress_ID, y = Root_Biomass, color = Stress_ID)) + 
  geom_boxplot(outlier.alpha = 0.4) + 
  geom_point(size = 2, alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept=71.40, linetype = "dotted") + 
    scale_color_manual(
    values = c(
        "Control" = "#e6194B",
        "MP" = "#3cb44b",
        "Drought" = "#ffe119",
        "Herbicide" = "#4363d8",
        "Antibiotic" = "#f58231",
        "Drought_MP" = "#42d4f4",
        "Antibiotic_MP" = "#f032e6",
        "Herbicide_MP" = "#469990",
        "Antibiotic_Herbicide" = "#9A6324",
        "Drought_Herbicide" = "#800000",
        "Antibiotic_Drought" = "#911eb4",
        "Antibiotic_Drought_Herbicide" = "#bfef45", 
        "Antibiotic_Drought_MP" = "#808000",
        "Antibiotic_Herbicide_MP" = "#000075",
        "Drought_Herbicide_MP" = "#a9a9a9", 
        "Antibiotic_Drought_Herbicide_MP" = "#000000"
    )) + 
  labs(y = "Root Biomass", x = "Treatment") + 
  theme(legend.position = "none")

ggsave("~/root_biomass_boxplot.pdf", root_biomass_boxplot)
```


```{r}
# combining the effect size plots for biomass 
# have to replot original plots first 
shoot_bio_multi_effects <- ggplot(plot_data, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                                  y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment", y = "Effect Size (95% CI)",
       title = "Shoot Biomass",
       color = "Interaction Type") + 
  theme(axis.text.x = element_blank()) + 
  theme(axis.title.x = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))

root_bio_multiplicative_effects <- ggplot(plot_data_rbiomass, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                                  y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment", y = "Effect Size (95% CI)",
       title = "Root Biomass",
       color = "Interaction") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(plot.title = element_text(hjust = 0.5))

biomass_eff_size_fig <- ggarrange(shoot_bio_multi_effects, root_bio_multiplicative_effects, 
                                  ncol = 1, nrow = 2, 
                                  common.legend = TRUE, legend = "right", 
                                  align = "hv")
)
ggsave("~/biomass_multi_eff_size_fig.pdf", biomass_eff_size_fig, width = 10, height = 14)
```

# Root Architecture #
## Checking for correlations among root architecutre traits ## 
```{r}
# Choose only the variables we want to test the linear correlations with 
vars <- c("Median.Number.of.Roots", "Maximum.Number.of.Roots", "Number.of.Root.Tips", 
          "Total.Root.Length.mm", "Depth.mm", "Maximum.Width.mm", "Width-to-Depth.Ratio", 
          "Network.Area.mm2", "Convex.Area.mm2", "Solidity", "Lower.Root.Area.mm2", 
          "Average.Diameter.mm", "Maximum.Diameter.mm", "Perimeter.mm", 
          "Volume.mm3", "Surface.Area.mm2", "Holes", "Average.Root.Orientation.deg", 
          "Shallow.Angle.Frequency", "Medium.Angle.Frequency", "Steep.Angle.Frequency", 
          "Root.Length.Diameter.Range.1.mm", "Root.Length.Diameter.Range.2.mm", 
          "Projected.Area.Diameter.Range.1.mm2", "Projected.Area.Diameter.Range.2.mm2", 
          "Surface.Area.Diameter.Range.1.mm2", "Surface.Area.Diameter.Range.2.mm2", 
          "Volume.Diameter.Range.1.mm3", "Volume.Diameter.Range.2.mm3")

sub_df <- mse1 %>%
  dplyr::select(all_of(vars)) %>%
  dplyr::select(where(is.numeric))

r2_mat <- cor(sub_df, use = "pairwise.complete.obs", method = "pearson")^2

r2_long <- as.data.frame(r2_mat) %>%
  tibble::rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "r2") %>%
  mutate(
    var1 = factor(var1, levels = vars),
    var2 = factor(var2, levels = vars),
    i = as.integer(var1),
    j = as.integer(var2)
  ) %>%
  filter(j <= i)  # Lower triangle including diagonal

root_arch_corr <- ggplot(r2_long, aes(x = var2, y = var1, fill = r2)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r2)), size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue", limits = c(0, 1)) +
  coord_equal() +
  scale_x_discrete(limits = vars) +
  scale_y_discrete(limits = rev(vars)) +
  labs(x = NULL, y = NULL, fill = expression(R^2),
       title = expression("Lower-triangle " * R^2 * " heat map")) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    # Rotate x-axis labels 90 degrees
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

ggsave("~/root_arch_coor.pdf", root_arch_corr, width = 14, height = 14)
```

## Multivariate summary of root traits across stress treatments ## 
```{r}
traits <- mse1_clean %>% dplyr::select(Depth.mm, Total.Root.Length.mm, Average.Diameter.mm, Average.Root.Orientation.deg, Volume.mm3, 
                                Convex.Area.mm2, Maximum.Number.of.Roots, Maximum.Width.mm, Maximum.Diameter.mm, Number.of.Root.Tips, 
                                Medium.Angle.Frequency, `Width-to-Depth.Ratio`) %>% 
  scale()

pca <- prcomp(traits, center = TRUE, scale. = FALSE)

# Proportion of variance explained by each PC
var_explained <- summary(pca)$importance["Proportion of Variance", ]

# Percent variance for PC1 and PC2
pc1_var <- var_explained[1] * 100
pc2_var <- var_explained[2] * 100

pc1_var
pc2_var

# plotting 
scores <- as.data.frame(pca$x[, 1:2])
scores$sample_id <- mse1_clean$ID
scores$stress_id <- mse1_clean$Stress_ID
scores$num_stress <- mse1_clean$Num_Stress

root_arch_pca <- ggplot(scores, aes(PC1, PC2, color = stress_id)) +
  geom_point(size = 5, alpha = 0.85) +
  labs(color = "Treatment") + 
  scale_color_manual(
    values = c(
        "Control" = "#e6194B",
        "MP" = "#3cb44b",
        "Drought" = "#ffe119",
        "Herbicide" = "#4363d8",
        "Antibiotic" = "#f58231",
        "Drought_MP" = "#42d4f4",
        "Antibiotic_MP" = "#f032e6",
        "Herbicide_MP" = "#469990",
        "Antibiotic_Herbicide" = "#9A6324",
        "Drought_Herbicide" = "#800000",
        "Antibiotic_Drought" = "#911eb4",
        "Antibiotic_Drought_Herbicide" = "#bfef45", 
        "Antibiotic_Drought_MP" = "#808000",
        "Antibiotic_Herbicide_MP" = "#000075",
        "Drought_Herbicide_MP" = "#a9a9a9", 
        "Antibiotic_Drought_Herbicide_MP" = "#000000"
    ))

ggsave("~/root_arch_pca.pdf", root_arch_pca, width = 11, height = 10)
```

## Root Depth ##
#### Calculating the multiplicative null model estimates and effect sizes 
```{r}
analyze_all_combos_depth <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_depth = mean(Depth.mm, na.rm = TRUE), 
      sd_depth = sd(Depth.mm, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_depth <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_depth <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_depth) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_depth
    Xc <-  control$mean_depth
    X_obs <- combo_data_depth$mean_depth
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_depth$sd_depth^2 / (combo_data_depth$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_depth^2 / (single_data$n * single_data$mean_depth^2)) +  # Singles
    (k - 1) * (control$sd_depth^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_depth <- map_dfr(combinations, ~calculate_effect_size_depth(.x, treatment_summaries))
  return(results_depth)
} 

# run it 
results_depth <- analyze_all_combos_depth(mse1)
print(results_depth)
```

#### Plotting the effect sizes
```{r}
plot_data_depth <- results_depth %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_depth_multiplicative_effects <- ggplot(plot_data_depth, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"), 
                                                                 y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment",
       title = "Depth",
       color = "Interaction Type") + 
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.x = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))
```

## Number of Root Tips ##
#### Calculating the multiplicative model estimates and effect sizes 
```{r}
analyze_all_combos_tips <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_tips = mean(Number.of.Root.Tips, na.rm = TRUE), 
      sd_tips = sd(Number.of.Root.Tips, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_tips <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_tips <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_tips) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_tips
    Xc <-  control$mean_tips
    X_obs <- combo_data_tips$mean_tips
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_tips$sd_tips^2 / (combo_data_tips$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_tips^2 / (single_data$n * single_data$mean_tips^2)) +  # Singles
    (k - 1) * (control$sd_tips^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_tips <- map_dfr(combinations, ~calculate_effect_size_tips(.x, treatment_summaries))
  return(results_tips)
} 

# run it 
results_tips <- analyze_all_combos_tips(mse1)
print(results_tips)
```

#### Plotting effect sizes
```{r}
plot_data_tips <- results_tips %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_tips_multi_effects <- ggplot(plot_data_tips, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"), 
                                                      y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment",
       title = "Number of Tips",
       color = "Interaction") + 
  theme(axis.title.x = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))
```

## Average Root Diameter ##
#### Calculating the multiplicative model estimates and effect sizes
```{r}
analyze_all_combos_diameter <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_diameter = mean(Average.Diameter.mm, na.rm = TRUE), 
      sd_diameter = sd(Average.Diameter.mm, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_diameter <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_diameter <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_diameter) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_diameter
    Xc <-  control$mean_diameter
    X_obs <- combo_data_diameter$mean_diameter
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_diameter$sd_diameter^2 / (combo_data_diameter$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_diameter^2 / (single_data$n * single_data$mean_diameter^2)) +  # Singles
    (k - 1) * (control$sd_diameter^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_diameter <- map_dfr(combinations, ~calculate_effect_size_diameter(.x, treatment_summaries))
  return(results_diameter)
} 

# run it 
results_diameter <- analyze_all_combos_diameter(mse1)
print(results_diameter )
```

#### Plotting effect sizes 
```{r}
plot_data_diameter <- results_diameter %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_diameter_multi_effects <- ggplot(plot_data_diameter, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                              y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment",
       title = "Avg. Diameter",
       color = "Interaction") + 
  theme(axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        axis.title.x = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Total Root Length ##
#### Calculating the multiplicative model estimates and effect sizes
```{r}
analyze_all_combos_length <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_length = mean(Total.Root.Length.mm, na.rm = TRUE), 
      sd_length = sd(Total.Root.Length.mm, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_length <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_length <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_length) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_length
    Xc <-  control$mean_length
    X_obs <- combo_data_length$mean_length
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_length$sd_length^2 / (combo_data_length$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_length^2 / (single_data$n * single_data$mean_length^2)) +  # Singles
    (k - 1) * (control$sd_length^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_length <- map_dfr(combinations, ~calculate_effect_size_length(.x, treatment_summaries))
  return(results_length)
} 

# run it 
results_length <- analyze_all_combos_length(mse1)
print(results_length)
```

#### Plotting effect sizes 
```{r}
plot_data_length <- results_length %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_length_multi_effects <- ggplot(plot_data_length, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                          y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment",
       title = "Total Root Length",
       color = "Interaction") + 
  theme(axis.title.x = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))
```

## Average Root Orientation ##
#### Calculating the multiplicative model estimates and effect sizes
```{r}
analyze_all_combos_orientation <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_orientation = mean(Average.Root.Orientation.deg, na.rm = TRUE), 
      sd_orientation = sd(Average.Root.Orientation.deg, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_orientation <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_orientation <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_orientation) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_orientation
    Xc <-  control$mean_orientation
    X_obs <- combo_data_orientation$mean_orientation
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_orientation$sd_orientation^2 / (combo_data_orientation$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_orientation^2 / (single_data$n * single_data$mean_orientation^2)) +  # Singles
    (k - 1) * (control$sd_orientation^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_orientation <- map_dfr(combinations, ~calculate_effect_size_orientation(.x, treatment_summaries))
  return(results_orientation)
} 

# run it 
results_orientation <- analyze_all_combos_orientation(mse1)
print(results_orientation)
```

#### Plotting effect sizes 
```{r}
plot_data_orientation <- results_orientation %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_orientation_multi_effects <- ggplot(plot_data_orientation, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                          y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment", y = "Effect Size (95% CI)",
       title = "Avg. Root Orientation",
       color = "Interaction") + 
  theme(axis.title.y = element_blank(), 
       axis.text.y = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))
```

## Maximum Number of Roots ## 
#### Calculating the multiplicative model estimates and effect sizes
```{r}
analyze_all_combos_maxroots <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_maxroots = mean(Maximum.Number.of.Roots, na.rm = TRUE), 
      sd_maxroots = sd(Maximum.Number.of.Roots, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_maxroots <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_maxroots <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_maxroots) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_maxroots
    Xc <-  control$mean_maxroots
    X_obs <- combo_data_maxroots$mean_maxroots
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_maxroots$sd_maxroots^2 / (combo_data_maxroots$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_maxroots^2 / (single_data$n * single_data$mean_maxroots^2)) +  # Singles
    (k - 1) * (control$sd_maxroots^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_maxroots <- map_dfr(combinations, ~calculate_effect_size_maxroots(.x, treatment_summaries))
  return(results_maxroots)
} 

# run it 
results_maxroots <- analyze_all_combos_maxroots(mse1)
print(results_maxroots)
```

#### Plotting effect sizes 
```{r}
plot_data_maxroots <- results_maxroots %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

max_root_multi_effects <- ggplot(plot_data_maxroots, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                          y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment", y = "Effect Size (95% CI)",
       title = "Max. Number of Roots",
       color = "Interaction") + 
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))
```

## Maximum Width ## 
#### Calculating the multiplicative model estimates and effect sizes
```{r}
analyze_all_combos_width <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_width = mean(Maximum.Width.mm, na.rm = TRUE), 
      sd_width = sd(Maximum.Width.mm, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_width <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_width <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_width) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_width
    Xc <-  control$mean_width
    X_obs <- combo_data_width$mean_width
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_width$sd_width^2 / (combo_data_width$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_width^2 / (single_data$n * single_data$mean_width^2)) +  # Singles
    (k - 1) * (control$sd_width^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_width <- map_dfr(combinations, ~calculate_effect_size_width(.x, treatment_summaries))
  return(results_width)
} 

# run it 
results_width <- analyze_all_combos_width(mse1)
print(results_width)
```

#### Plotting effect sizes 
```{r}
plot_data_width <- results_width %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_width_multi_effects <- ggplot(plot_data_width, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                          y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment",
       title = "Max. Width",
       color = "Interaction") + 
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.x = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))
```

## Volume ## 
#### Calculating the multiplicative model estimates and effect sizes
```{r}
analyze_all_combos_volume <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_volume = mean(Volume.mm3, na.rm = TRUE), 
      sd_volume = sd(Volume.mm3, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_volume <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_volume <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_volume) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_volume
    Xc <-  control$mean_volume
    X_obs <- combo_data_volume$mean_volume
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_volume$sd_volume^2 / (combo_data_volume$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_volume^2 / (single_data$n * single_data$mean_volume^2)) +  # Singles
    (k - 1) * (control$sd_volume^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_volume <- map_dfr(combinations, ~calculate_effect_size_volume(.x, treatment_summaries))
  return(results_volume)
} 

# run it 
results_volume <- analyze_all_combos_volume(mse1)
print(results_volume)
```

#### Plotting effect sizes 
```{r}
plot_data_volume <- results_volume %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_volume_multi_effects <- ggplot(plot_data_volume, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                          y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment",
       title = "Volume",
       color = "Interaction") +
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.x = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))
```

## Width-to-depth ratio ## 
#### Calculating the multiplicative model estimates and effect sizes
```{r}
analyze_all_combos_wtd <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_wtd = mean(`Width-to-Depth.Ratio`, na.rm = TRUE), 
      sd_wtd = sd(`Width-to-Depth.Ratio`, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_wtd <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_wtd <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_wtd) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_wtd
    Xc <-  control$mean_wtd
    X_obs <- combo_data_wtd$mean_wtd
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_wtd$sd_wtd^2 / (combo_data_wtd$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_wtd^2 / (single_data$n * single_data$mean_wtd^2)) +  # Singles
    (k - 1) * (control$sd_wtd^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_wtd <- map_dfr(combinations, ~calculate_effect_size_wtd(.x, treatment_summaries))
  return(results_wtd)
} 

# run it 
results_wtd <- analyze_all_combos_wtd(mse1)
print(results_wtd)
```

#### Plotting effect sizes 
```{r}
plot_data_wtd <- results_volume %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_wtd_multi_effects <- ggplot(plot_data_wtd, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                          y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment", y = "Effect Size (95% CI)",
       title = "Width:Depth",
       color = "Interaction") + 
  theme(plot.title = element_text(hjust = 0.5))
#+ theme(axis.title.y = element_blank(), 
#        axis.text.y = element_blank())
```

## Convex Area ## 
#### Calculating the multiplicative model estimates and effect sizes
```{r}
analyze_all_combos_convex <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_convex = mean(Convex.Area.mm2, na.rm = TRUE), 
      sd_convex = sd(Convex.Area.mm2, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_convex <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_convex <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_convex) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_convex
    Xc <-  control$mean_convex
    X_obs <- combo_data_convex$mean_convex
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_convex$sd_convex^2 / (combo_data_convex$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_convex^2 / (single_data$n * single_data$mean_convex^2)) +  # Singles
    (k - 1) * (control$sd_convex^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_convex <- map_dfr(combinations, ~calculate_effect_size_convex(.x, treatment_summaries))
  return(results_convex)
} 

# run it 
results_convex <- analyze_all_combos_convex(mse1)
print(results_convex)
```

#### Plotting effect sizes 
```{r}
plot_data_convex <- results_convex %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_convex_multi_effects <- ggplot(plot_data_convex, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                          y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment", y = "Effect Size (95% CI)",
       title = "Convex Area",
       color = "Interaction") +
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))
```

## Maximum Diameter ## 
#### Calculating the multiplicative model estimates and effect sizes
```{r}
analyze_all_combos_maxdiameter <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_maxdiameter = mean(Maximum.Diameter.mm, na.rm = TRUE), 
      sd_maxdiameter = sd(Maximum.Diameter.mm, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_maxdiameter <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_maxdiameter <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_maxdiameter) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_maxdiameter
    Xc <-  control$mean_maxdiameter
    X_obs <- combo_data_maxdiameter$mean_maxdiameter
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_maxdiameter$sd_maxdiameter^2 / (combo_data_maxdiameter$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_maxdiameter^2 / (single_data$n * single_data$mean_maxdiameter^2)) +  # Singles
    (k - 1) * (control$sd_maxdiameter^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_maxdiameter <- map_dfr(combinations, ~calculate_effect_size_maxdiameter(.x, treatment_summaries))
  return(results_maxdiameter)
} 

# run it 
results_maxdiameter <- analyze_all_combos_maxdiameter(mse1)
print(results_maxdiameter)
```

#### Plotting effect sizes 
```{r}
plot_data_maxdiameter <- results_maxdiameter %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_maxdiameter_multi_effects <- ggplot(plot_data_maxdiameter, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                          y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment",
       title = "Max. Diameter",
       color = "Interaction") +
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.x = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))
```

## Medium Angle Frequency ## 
#### Calculating the multiplicative model estimates and effect sizes
```{r}
analyze_all_combos_medangle <- function(mse1) {
  
  # extract all treatment summaries
  treatment_summaries <- mse1 %>% 
    group_by(Stress_ID) %>% 
    summarize( 
      mean_medangle = mean(Medium.Angle.Frequency, na.rm = TRUE), 
      sd_medangle = sd(Medium.Angle.Frequency, na.rm = TRUE), 
      n = n(), 
      .groups = "drop")
  # define all combinations to test 
  combinations <- list(
    #pairwise 
    list(singles = c("MP", "Drought"), combo = "Drought_MP"), 
    list(singles = c("MP", "Herbicide"), combo = "Herbicide_MP"), 
    list(singles = c("MP", "Antibiotic"), combo = "Antibiotic_MP"), 
    list(singles = c("Drought", "Herbicide"), combo = "Drought_Herbicide"),
    list(singles = c("Drought", "Antibiotic"), combo = "Antibiotic_Drought"),
    list(singles = c("Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide"),
    #three-way 
    list(singles = c("MP", "Drought", "Herbicide"), combo = "Drought_Herbicide_MP"), 
    list(singles = c("MP", "Drought", "Antibiotic"), combo = "Antibiotic_Drought_MP"), 
    list(singles = c("MP", "Herbicide", "Antibiotic"), combo = "Antibiotic_Herbicide_MP"), 
    list(singles = c("Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide"),
    #four-way 
    list(singles = c("MP", "Drought", "Herbicide", "Antibiotic"), combo = "Antibiotic_Drought_Herbicide_MP")
  )
  calculate_effect_size_medangle <- function(combo_info, summaries) {
    control <- summaries %>% filter(Stress_ID == "Control")
    combo_data_medangle <- summaries %>% filter(Stress_ID == combo_info$combo)
    
    if(nrow(control) == 0 | nrow(combo_data_medangle) == 0) return(NULL)
    
    # get single stressor data 
    single_data <- map_dfr(combo_info$singles, ~{
      summaries %>% filter(Stress_ID == .x)
    })
    
    if(nrow(single_data) == 0) return(NULL)
    
    # multiplicative expectation 
    single_means <- single_data$mean_medangle
    Xc <-  control$mean_medangle
    X_obs <- combo_data_medangle$mean_medangle
    
    X_exp <- prod(single_means) / Xc^(length(single_means) -1)
    
    # simple effect size 
    ES <- log(X_obs) - log(X_exp)
    
    # VARIANCE using delta method (all treatments contribute)
  k <- length(single_means)  # number of single stressors
  
  var_ES <- (
    combo_data_medangle$sd_medangle^2 / (combo_data_medangle$n * X_obs^2) +           # Observed combo
    sum(single_data$sd_medangle^2 / (single_data$n * single_data$mean_medangle^2)) +  # Singles
    (k - 1) * (control$sd_medangle^2 / (control$n * Xc^2))          # Control (raised to k-1)
  )
  
  SE <- sqrt(var_ES)
  CI_lower <- ES - 1.96 * SE
  CI_upper <- ES + 1.96 * SE
  
  # Classification
  classification <- case_when(
    CI_lower > 0 ~ "Synergistic",
    CI_upper < 0 ~ "Antagonistic",
    CI_lower <= 0 & CI_upper >= 0 ~ "Null",
    TRUE ~ "Reversal"
  )
  
    data.frame(
      combination = combo_info$combo,
      singles_used = paste(combo_info$singles, collapse = "_"), 
      observed = X_obs, 
      expected = X_exp, 
      effect_size = ES, 
      SE = SE, 
      CI_lower = CI_lower, 
      CI_upper = CI_upper,
      classification = classification,
      stringsAsFactors = FALSE
    )
  }
  
  # apply to all combos 
  results_medangle <- map_dfr(combinations, ~calculate_effect_size_medangle(.x, treatment_summaries))
  return(results_medangle)
} 

# run it 
results_medangle <- analyze_all_combos_medangle(mse1)
print(results_medangle)
```

#### Plotting effect sizes 
```{r}
plot_data_medangle <- results_medangle %>%
  mutate(n_stressors = str_count(singles_used, "_") + 2)

root_medangle_multi_effects <- ggplot(plot_data_medangle, aes(x = fct_relevel(combination,
                                                                  "Antibiotic_Drought", 
                                                                  "Antibiotic_Herbicide", 
                                                                  "Antibiotic_MP",
                                                                  "Drought_Herbicide", 
                                                                  "Drought_MP", 
                                                                  "Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide", 
                                                                  "Antibiotic_Drought_MP", 
                                                                  "Antibiotic_Herbicide_MP", 
                                                                  "Drought_Herbicide_MP", 
                                                                  "Antibiotic_Drought_Herbicide_MP"),
                                                          y = effect_size, color = classification)) +
  geom_hline(yintercept = 0, color = "black", size = 1, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.9) +
  coord_flip() +
  scale_color_manual(values = c("Antagonistic" = "#4363d8", "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) +
  labs(x = "Stress Treatment",
       title = "Medium Angle Freq.",
       color = "Interaction") +
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.x = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))
```

# Conclusions 
```{r}
# combining data frames so I can see how many of each interaction classification per response variable in one plot 
df_list <- list(results, results_rbiomass, results_depth, results_diameter, results_tips, results_length, results_maxdiameter, results_maxroots, results_convex, results_medangle, 
                results_orientation, results_volume, results_width, results_wtd) 
names(df_list) <- c("shoot_biomass", "root_biomass", "root_depth", "root_average_diameter", "root_tips", "total_root_length", "root_maximum_diameter", 
                    "maximum_number_of_roots", "root_convex_area", "root_medium_angle_frequency", "average_root_orientation", "root_volume", "root_width",
                    "root_width_to_depth_ratio")

combined_df <- bind_rows(df_list, .id = "Response")

# Define a custom order of treatments for the x axis 
treatment_order <- factor(c("Control", "MP", "Drought", "Herbicide", "Antibiotic",
                          "Drought_MP", "Herbicide_MP", "Antibiotic_MP", 
                          "Drought_Herbicide", "Antibiotic_Drought", "Antibiotic_Herbicide",
                          "Drought_Herbicide_MP", "Antibiotic_Drought_MP", 
                          "Antibiotic_Herbicide_MP", "Antibiotic_Drought_Herbicide", 
                          "Antibiotic_Drought_Herbicide_MP"), 
                         levels = c("Control", "MP", "Drought", "Herbicide", "Antibiotic",
                                   "Drought_MP", "Herbicide_MP", "Antibiotic_MP", 
                                   "Drought_Herbicide", "Antibiotic_Drought", "Antibiotic_Herbicide",
                                   "Drought_Herbicide_MP", "Antibiotic_Drought_MP", 
                                   "Antibiotic_Herbicide_MP", "Antibiotic_Drought_Herbicide", 
                                  "Antibiotic_Drought_Herbicide_MP"))

combined_df$combination <- factor(combined_df$combination, levels = levels(treatment_order))

# separate by category of response variable 
## (i.e., root + shoot biomass = Biomass; length, tips, surface area, diameter, depth, etc. = Root Architecture)

# add a category column to the df 
combined_df <- combined_df %>% 
  mutate(Category = case_when(
    Response %in% c("shoot_biomass", "root_biomass") ~ "Biomass", 
    Response %in% c("root_depth", "root_average_diameter", "root_tips", "total_root_length", "root_maximum_diameter", 
                    "maximum_number_of_roots", "root_convex_area", "root_medium_angle_frequency", "average_root_orientation", "root_volume", "root_width",
                    "root_width_to_depth_ratio") ~ "Root Architecture"
  ))

# subsetting the dataframe to separate biomass from architecture 
biomass_df <- combined_df %>% 
  filter(Response %in% c("shoot_biomass", "root_biomass"))

architecture_df <- combined_df %>% 
  filter(Response %in% c("root_tips", "root_depth", "root_average_diameter", "root_maximum_diameter", "total_root_length",
                    "root_medium_angle_frequency", "root_volume", "root_width", "root_width_to_depth_ratio", "average_root_orientation",
                    "root_convex_area", "maximum_number_of_roots"))

# now plot separately 
p_biomass <- ggplot(biomass_df, aes(x = Response, fill = classification)) + 
  geom_bar(position = "fill") + 
  labs(x = "Response", y = "Proportion of Interactions", 
       fill = "Interaction Type") + 
  scale_fill_manual(values = c("Antagonistic" = "#4363d8", 
                              "Synergistic" = "#e6194B", 
                              "Null" = "gray50")) + 
  scale_x_discrete(labels = c("Root Biomass", "Shoot Biomass")) + 
  theme(axis.text.x = element_text(angle =45, hjust =1))

ggsave("~/prop_biomass_classification.pdf", p_biomass, height = 4, width = 3)

p_arch <- ggplot(architecture_df, aes(x = Response, fill = classification)) + 
  geom_bar(position = "fill") + 
  labs(x = "Response", y = "Proportion of Interactions", 
       fill = "Interaction Type") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_fill_manual(values = c("Antagonistic" = "#4363d8", 
                              "Synergistic" = "#e6194B", 
                              "Null" = "gray50")) + 
  scale_x_discrete(labels = c("Avg. Root Orientation", "Max. Number of Roots", "Avg. Diameter", 
                              "Convex Area", "Depth", "Max. Diameter", "Medium Angle Freq.", "Number of Tips", 
                              "Volume", "Max. Width", "Width:Depth", "Total Root Length"
  ))

ggsave("~/prop_architecture_classifications.pdf", p_arch, width = 6, height =4)

# plot stress ID by biomass and arch 
p_biomass_stress <- ggplot(biomass_df, aes(x = combination, fill = classification)) + 
  geom_bar(position = "fill") + 
  labs(x = "Stress Treatment", y = "Proportion of Interactions", fill = "Interaction Type") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_fill_manual(values = c("Antagonistic" = "#4363d8", 
                               "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) + 
  scale_x_discrete(labels = c("Drought + MP", "Herbicide + MP", "Antibiotic + MP", 
                              "Drought + Herbicide", "Antibiotic + Drought", "Antibiotic + Herbicide", 
                              "Drought + Herbicide + MP", "Antibiotic + Drought + MP", 
                              "Antibiotic + Herbicide + MP", "Antibiotic + Drought + Herbicide", 
                              "Antibiotic + Drought + Herbicide + MP"))

ggsave("~/prop_biomass_by_stress.pdf", p_biomass_stress, width = 5, height = 4)

p_arch_stress <- ggplot(architecture_df, aes(x = combination, fill = classification)) + 
  geom_bar(position = "fill") + 
  labs(x = "Stress Treatment", y = "Proportion of Interactions", fill = "Interaction Type") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_fill_manual(values = c("Antagonistic" = "#4363d8", 
                               "Synergistic" = "#e6194B", 
                               "Null" = "gray50")) + 
    scale_x_discrete(labels = c("Drought + MP", "Herbicide + MP", "Antibiotic + MP", 
                              "Drought + Herbicide", "Antibiotic + Drought", "Antibiotic + Herbicide", 
                              "Drought + Herbicide + MP", "Antibiotic + Drought + MP", 
                              "Antibiotic + Herbicide + MP", "Antibiotic + Drought + Herbicide", 
                              "Antibiotic + Drought + Herbicide + MP"))

ggsave("~/prop_arch_by_stress.pdf", p_arch_stress, width = 5, height = 4)
```


```{r}
# combining the root architecture effect size plots 
arch_eff_size_fig <- ggarrange(root_tips_multi_effects, root_depth_multiplicative_effects,
                              root_diameter_multi_effects, root_maxdiameter_multi_effects,
                              root_length_multi_effects,
                              root_medangle_multi_effects, root_volume_multi_effects, 
                              root_width_multi_effects, root_wtd_multi_effects,
                              root_orientation_multi_effects, 
                              root_convex_multi_effects, max_root_multi_effects,
                              ncol = 4, nrow = 3, 
                              common.legend = TRUE, 
                              align = "hv")

ggsave("~/arch_eff_size_full_fig.pdf", arch_eff_size_fig, width = 18, height = 14)
```
