---
title: "MSE1_Drought"
output: html_document
date: "2026-05-26"
---
Drought treatment in MSE1 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages 
library(ggplot2)
library(readxl)
library(dplyr)
library(nlme)
library(lme4)
library(lmerTest)

# set theme 
set_theme(theme_bw())
```

```{r}
# load data 
x <- read_xlsx("~/Desktop/FDA_Lab/Experiment1/manuscript/data/SoilProbe.xlsx")

x$Stress_ID <- as.factor(x$Stress_ID)
x$Drought <- as.factor(x$Drought)
```
In the experiment, we only included soil probes in one pot per block (and note that these are not true "blocks". This was a completely random design and I call the racks that the pots were stacked in "blocks"). Probes were also placed so that each unique treatment had one probe measuring soil moisture. I collected soil moisture measurements every two days immediately post-watering. For treatments that were droughted, I collected soil probe readings after watering the appropriate treatments. 

```{r}
# plot soil moisture across all probes and the full experiment 
soilmoisture <- ggplot(x, aes(x = Time, y = Probe_Reading, color = Drought, group = ID)) + 
  geom_line() + 
  labs(x = "Time", 
       y = "Soil Moisture", 
       color = "Drought Treatment") + 
  geom_vline(xintercept = 16, 
             linetype = "dashed") + 
    scale_color_manual(values = c("Presence" = "#e6194B", 
                                "Absence" = "#4363d8"))

ggsave("~/Desktop/FDA_Lab/Experiment1/manuscript/plots/soilmoisturefull.pdf")
```
Why is there a huge drop off between time point 20 and 21 regardless of drought treatment? Timepoint 21 was between waterings, whereas all other measurements were immediately post-watering 

```{r}
# soil probe reading at only the last time point (T = 21)
lasttime <- x[x$Time == 21, ]

lasttime$Stress_ID <- factor(lasttime$Stress_ID, 
                            levels = c("Control", "MP", "Herbicide", "Antibiotic",
                                       "Herbicide_MP", "Antibiotic_MP", "Antibiotic_Herbicide", 
                                       "Antibiotic_Herbicide_MP", 
                                       "Drought", "Drought_MP", "Drought_Herbicide", 
                                       "Antibiotic_Drought", "Drought_Herbicide_MP", 
                                       "Antibiotic_Drought_MP", "Antibiotic_Drought_Herbicide", 
                                       "Antibiotic_Drought_Herbicide_MP"))

soilprobelasttime <- ggplot(lasttime, aes(x = Stress_ID, y = Probe_Reading, color = Drought)) + 
  geom_point(size = 3) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_color_manual(values = c("Presence" = "#e6194B", 
                                "Absence" = "#4363d8")) + 
  labs(x = "Treatment", 
       y = "Probe Reading")

ggsave("~/Desktop/FDA_Lab/Experiment1/manuscript/plots/soilprobelasttime.pdf", soilprobelasttime)

# try boxplot 
soilprobelasttimeboxplot <- ggplot(lasttime, aes(x = Drought, y = Probe_Reading, color = Drought)) + 
  geom_boxplot() + 
  geom_point(alpha = 0.5, size = 2) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_color_manual(values = c("Presence" = "#e6194B", 
                                "Absence" = "#4363d8")) + 
  labs(x = "Treatment", 
       y = "Probe Reading")

ggsave("~/Desktop/FDA_Lab/Experiment1/manuscript/plots/soilprobelasttimeboxplot.pdf", soilprobelasttimeboxplot)


# soil probe reading at the penultimate time point (T = 20)
penultimate <- x[x$Time == 20, ]

penultimate$Stress_ID <- factor(penultimate$Stress_ID, 
                                levels = c("Control", "MP", "Herbicide", "Antibiotic",
                                       "Herbicide_MP", "Antibiotic_MP", "Antibiotic_Herbicide", 
                                       "Antibiotic_Herbicide_MP", 
                                       "Drought", "Drought_MP", "Drought_Herbicide", 
                                       "Antibiotic_Drought", "Drought_Herbicide_MP", 
                                       "Antibiotic_Drought_MP", "Antibiotic_Drought_Herbicide", 
                                       "Antibiotic_Drought_Herbicide_MP"))

soilprobepenultimate <- ggplot(penultimate, aes(x = Stress_ID, y = Probe_Reading, color = Drought)) + 
  geom_point(size = 3) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_color_manual(values = c("Presence" = "#e6194B", 
                                "Absence" = "#4363d8")) + 
  labs(x = "Treatment", 
       y = "Probe Reading")

ggsave("~/Desktop/FDA_Lab/Experiment1/manuscript/plots/soilprobepenultimate.pdf", soilprobepenultimate)

# boxplot 
soilprobepenultimateboxplot <- ggplot(penultimate, aes(x = Drought, y = Probe_Reading, color = Drought)) + 
  geom_boxplot() + 
  geom_point(alpha = 0.5, size = 2) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_color_manual(values = c("Presence" = "#e6194B", 
                                "Absence" = "#4363d8")) + 
  labs(x = "Treatment", 
       y = "Probe Reading")

ggsave("~/Desktop/FDA_Lab/Experiment1/manuscript/plots/soilprobepenultimateboxplot.pdf", soilprobepenultimateboxplot)

# comparing drought vs control for the whole experiment
ggplot(x, aes(x = Drought, y = Probe_Reading, color = Drought)) + 
  geom_boxplot() + 
  geom_point(alpha = 0.5, size = 2) + 
  scale_color_manual(values = c("Presence" = "#e6194B", 
                                "Absence" = "#4363d8")) + 
  labs(x = "Treatment", 
       y = "Probe Reading")

# comparing drought vs control for just the drought portion of the experiment
soil_drought <- x %>%
  filter(Time %in% 16:21)

droughtvscontrol <- ggplot(soil_drought, aes(x = Drought, y = Probe_Reading, color = Drought)) + 
  geom_boxplot() + 
  geom_point(alpha = 0.5, size = 2) + 
  scale_color_manual(values = c("Presence" = "#e6194B", 
                                "Absence" = "#4363d8")) + 
  labs(x = "Treatment", 
       y = "Probe Reading")

ggsave("~/Desktop/FDA_Lab/Experiment1/manuscript/plots/droughtvscontrol.pdf", droughtvscontrol)
```


```{r}
# to show that the last time point drought pots are drier: 
lasttime
t.test(Probe_Reading ~ Drought, data = lasttime) # provides a mean difference and p-value for drought vs no drought at this time point

penultimate
t.test(Probe_Reading ~ Drought, data = penultimate)

soil_drought
t.test(Probe_Reading ~ Drought, data = soil_drought)
##
```
