---
title: "MSE1_dada2_processing_v3"
output: html_document
date: "2025-12-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1: Initial Library
```{r, include=FALSE}
library(rlang)
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("phyloseq")

#BiocManager::install("dada2")
library(dada2); packageVersion("dada2") #[1] '1.32.0'
library(Biostrings)
library(ShortRead)
library(tidyverse)
library(phyloseq)
library(ALDEx2)
library(vegan)
library(readxl)

#BiocManager::install("genefilter")
library(genefilter)
```

Step 2: Define the following path variable so that it points to the extracted directory on your local computer
```{r}
path <- "./raw_reads/"
#Check path
list.files(path)
path
```

Step 3: Read the names of fastq files 
```{r}
# Forward and reverse fastq filenames have format: PROJECT_RH_SAMPLEID_16S_R1_001.fastq and PROJECT_RH_SAMPLEID_16S_R2_001.fastq
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# Extract sample names, assuming filenames have format: PROJECT_RH_SAMPLEID_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 3)

# Inspect read quality profiles 
#The sequencing is 2x300 bp
plotQualityProfile(fnFs[1:2]) # not corrupted
plotQualityProfile(fnRs[1:2]) # not corrupted
```

```{r}
path_g.p <- file.path(path, "cutadapt_primers")   # <- where the *_cut.fastq live

fnFs.g.p <- sort(list.files(path_g.p, pattern = "_R1_cut2\\.fastq$", full.names = TRUE))
fnRs.g.p <- sort(list.files(path_g.p, pattern = "_R2_cut2\\.fastq$", full.names = TRUE))
```

Step 4: Inspect read quality profiles
```{r}
#The sequencing is 2x300 bp
plotQualityProfile(fnFs.g.p[2:3]) # 250
plotQualityProfile(fnRs.g.p[2:3]) 
```

Step 5: Filter and Trim 
```{r} 
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq"))

names(filtFs) <- sample.names
names(filtRs) <- sample.names

out <- filterAndTrim(fnFs.g.p, filtFs, fnRs.g.p, filtRs, #use the primers removed samples
                     trimLeft = c(0, 0), #primers removed cutadapt
                     truncLen=c(250, 240),
              maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=TRUE)

#obs: We’ll use standard filtering parameters: maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2. The maxEE parameter sets the maximum number of “expected errors” allowed in a read, which is a better filter than simply averaging quality scores

head(out)
save.image("./dada2_env_out.RData")
```

Step 6: Learn error rates 
https://github.com/ErnakovichLab/dada2_ernakovichlab#learn-the-error-rates
Traditional: Option 0
```{r}
#traditional
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
plotErrors(errF, nominalQ=TRUE)
#plotErrors(errR, nominalQ=TRUE)
```

The error rates don't look good; to deal with this, I am trying 4 options from the linked github pages and will choose the option that optimizes the error rates. 

Option 1 from JacobRPrice alter loess arguments (weights and span and enforce monotonicity)
benjjneb/dada2#1307
```{r}
library(magrittr)
library(dplyr)
loessErrfun_mod1 <- function(trans) {
  qq <- as.numeric(colnames(trans))
  est <- matrix(0, nrow=0, ncol=length(qq))
  for(nti in c("A","C","G","T")) {
    for(ntj in c("A","C","G","T")) {
      if(nti != ntj) {
        errs <- trans[paste0(nti,"2",ntj),]
        tot <- colSums(trans[paste0(nti,"2",c("A","C","G","T")),])
        rlogp <- log10((errs+1)/tot)  # 1 psuedocount for each err, but if tot=0 will give NA
        rlogp[is.infinite(rlogp)] <- NA
        df <- data.frame(q=qq, errs=errs, tot=tot, rlogp=rlogp)
        
        # original
        # ###! mod.lo <- loess(rlogp ~ q, df, weights=errs) ###!
        # mod.lo <- loess(rlogp ~ q, df, weights=tot) ###!
        # #        mod.lo <- loess(rlogp ~ q, df)
        
        # Gulliem Salazar's solution
        # https://github.com/benjjneb/dada2/issues/938
        mod.lo <- loess(rlogp ~ q, df, weights = log10(tot),span = 2)
        
        pred <- predict(mod.lo, qq)
        maxrli <- max(which(!is.na(pred)))
        minrli <- min(which(!is.na(pred)))
        pred[seq_along(pred)>maxrli] <- pred[[maxrli]]
        pred[seq_along(pred)<minrli] <- pred[[minrli]]
        est <- rbind(est, 10^pred)
      } # if(nti != ntj)
    } # for(ntj in c("A","C","G","T"))
  } # for(nti in c("A","C","G","T"))
  
  # HACKY
  MAX_ERROR_RATE <- 0.25
  MIN_ERROR_RATE <- 1e-7
  est[est>MAX_ERROR_RATE] <- MAX_ERROR_RATE
  est[est<MIN_ERROR_RATE] <- MIN_ERROR_RATE
  
  # enforce monotonicity
  # https://github.com/benjjneb/dada2/issues/791
  estorig <- est
  est <- est %>%
    data.frame() %>%
    mutate_all(funs(case_when(. < X40 ~ X40,
                              . >= X40 ~ .))) %>% as.matrix()
  rownames(est) <- rownames(estorig)
  colnames(est) <- colnames(estorig)
  
  # Expand the err matrix with the self-transition probs
  err <- rbind(1-colSums(est[1:3,]), est[1:3,],
               est[4,], 1-colSums(est[4:6,]), est[5:6,],
               est[7:8,], 1-colSums(est[7:9,]), est[9,],
               est[10:12,], 1-colSums(est[10:12,]))
  rownames(err) <- paste0(rep(c("A","C","G","T"), each=4), "2", c("A","C","G","T"))
  colnames(err) <- colnames(trans)
  # Return
  return(err)
}

# check what this looks like
errF_1 <- learnErrors(
  filtFs,
  multithread = TRUE,
  nbases = 1e8,
  errorEstimationFunction = loessErrfun_mod1,
  verbose = TRUE
)
## 108604125 total bases in 482685 reads from 1 samples will be used for learning the error rates.
## Initializing error rates to maximum possible estimate.
## selfConsist step 1 .
##    selfConsist step 2
##    selfConsist step 3
##    selfConsist step 4
##    selfConsist step 5
##    selfConsist step 6
##    selfConsist step 7
## Convergence after  7  rounds.
errR_1 <- learnErrors(
  filtRs,
  multithread = TRUE,
  nbases = 1e8,
  errorEstimationFunction = loessErrfun_mod1,
  verbose = TRUE
)
## 106190700 total bases in 482685 reads from 1 samples will be used for learning the error rates.
## Initializing error rates to maximum possible estimate.
## selfConsist step 1 .
##    selfConsist step 2
##    selfConsist step 3
##    selfConsist step 4
##    selfConsist step 5
##    selfConsist step 6
##    selfConsist step 7
##    selfConsist step 8
##    selfConsist step 9
##    selfConsist step 10
```

Option 2 enforce monotonicity only.
Originally recommended in: benjjneb/dada2#791
```{r}
loessErrfun_mod2 <- function(trans) {
  qq <- as.numeric(colnames(trans))
  est <- matrix(0, nrow=0, ncol=length(qq))
  for(nti in c("A","C","G","T")) {
    for(ntj in c("A","C","G","T")) {
      if(nti != ntj) {
        errs <- trans[paste0(nti,"2",ntj),]
        tot <- colSums(trans[paste0(nti,"2",c("A","C","G","T")),])
        rlogp <- log10((errs+1)/tot)  # 1 psuedocount for each err, but if tot=0 will give NA
        rlogp[is.infinite(rlogp)] <- NA
        df <- data.frame(q=qq, errs=errs, tot=tot, rlogp=rlogp)
        
        # original
        # ###! mod.lo <- loess(rlogp ~ q, df, weights=errs) ###!
        mod.lo <- loess(rlogp ~ q, df, weights=tot) ###!
        # #        mod.lo <- loess(rlogp ~ q, df)
        
        # Gulliem Salazar's solution
        # https://github.com/benjjneb/dada2/issues/938
        # mod.lo <- loess(rlogp ~ q, df, weights = log10(tot),span = 2)
        
        pred <- predict(mod.lo, qq)
        maxrli <- max(which(!is.na(pred)))
        minrli <- min(which(!is.na(pred)))
        pred[seq_along(pred)>maxrli] <- pred[[maxrli]]
        pred[seq_along(pred)<minrli] <- pred[[minrli]]
        est <- rbind(est, 10^pred)
      } # if(nti != ntj)
    } # for(ntj in c("A","C","G","T"))
  } # for(nti in c("A","C","G","T"))
  
  # HACKY
  MAX_ERROR_RATE <- 0.25
  MIN_ERROR_RATE <- 1e-7
  est[est>MAX_ERROR_RATE] <- MAX_ERROR_RATE
  est[est<MIN_ERROR_RATE] <- MIN_ERROR_RATE
  
  # enforce monotonicity
  # https://github.com/benjjneb/dada2/issues/791
  estorig <- est
  est <- est %>%
    data.frame() %>%
    mutate_all(funs(case_when(. < X40 ~ X40,
                              . >= X40 ~ .))) %>% as.matrix()
  rownames(est) <- rownames(estorig)
  colnames(est) <- colnames(estorig)
  
  # Expand the err matrix with the self-transition probs
  err <- rbind(1-colSums(est[1:3,]), est[1:3,],
               est[4,], 1-colSums(est[4:6,]), est[5:6,],
               est[7:8,], 1-colSums(est[7:9,]), est[9,],
               est[10:12,], 1-colSums(est[10:12,]))
  rownames(err) <- paste0(rep(c("A","C","G","T"), each=4), "2", c("A","C","G","T"))
  colnames(err) <- colnames(trans)
  # Return
  return(err)
}


# check what this looks like
errF_2 <- learnErrors(
  filtFs,
  multithread = TRUE,
  nbases = 1e8,
  errorEstimationFunction = loessErrfun_mod2,
  verbose = TRUE
)
## 108604125 total bases in 482685 reads from 1 samples will be used for learning the error rates.
## Initializing error rates to maximum possible estimate.
## selfConsist step 1 .
##    selfConsist step 2
##    selfConsist step 3
##    selfConsist step 4
##    selfConsist step 5
##    selfConsist step 6
##    selfConsist step 7
## Convergence after  7  rounds.

errR_2 <- learnErrors(
  filtRs,
  multithread = TRUE,
  nbases = 1e8,
  errorEstimationFunction = loessErrfun_mod2,
  verbose = TRUE
)
## 106190700 total bases in 482685 reads from 1 samples will be used for learning the error rates.
## Initializing error rates to maximum possible estimate.
## selfConsist step 1 .
##    selfConsist step 2
##    selfConsist step 3
##    selfConsist step 4
##    selfConsist step 5
##    selfConsist step 6
## Convergence after  6  rounds.
```
Option 3 alter loess function (weights only) and enforce monotonicity
From JacobRPrice benjjneb/dada2#1307
```{r}
loessErrfun_mod3 <- function(trans) {
  qq <- as.numeric(colnames(trans))
  est <- matrix(0, nrow=0, ncol=length(qq))
  for(nti in c("A","C","G","T")) {
    for(ntj in c("A","C","G","T")) {
      if(nti != ntj) {
        errs <- trans[paste0(nti,"2",ntj),]
        tot <- colSums(trans[paste0(nti,"2",c("A","C","G","T")),])
        rlogp <- log10((errs+1)/tot)  # 1 psuedocount for each err, but if tot=0 will give NA
        rlogp[is.infinite(rlogp)] <- NA
        df <- data.frame(q=qq, errs=errs, tot=tot, rlogp=rlogp)
        
        # original
        # ###! mod.lo <- loess(rlogp ~ q, df, weights=errs) ###!
        # mod.lo <- loess(rlogp ~ q, df, weights=tot) ###!
        # #        mod.lo <- loess(rlogp ~ q, df)
        
        # Gulliem Salazar's solution
        # https://github.com/benjjneb/dada2/issues/938
        # mod.lo <- loess(rlogp ~ q, df, weights = log10(tot),span = 2)
        
        # only change the weights
        mod.lo <- loess(rlogp ~ q, df, weights = log10(tot))
        
        pred <- predict(mod.lo, qq)
        maxrli <- max(which(!is.na(pred)))
        minrli <- min(which(!is.na(pred)))
        pred[seq_along(pred)>maxrli] <- pred[[maxrli]]
        pred[seq_along(pred)<minrli] <- pred[[minrli]]
        est <- rbind(est, 10^pred)
      } # if(nti != ntj)
    } # for(ntj in c("A","C","G","T"))
  } # for(nti in c("A","C","G","T"))
  
  # HACKY
  MAX_ERROR_RATE <- 0.25
  MIN_ERROR_RATE <- 1e-7
  est[est>MAX_ERROR_RATE] <- MAX_ERROR_RATE
  est[est<MIN_ERROR_RATE] <- MIN_ERROR_RATE
  
  # enforce monotonicity
  # https://github.com/benjjneb/dada2/issues/791
  estorig <- est
  est <- est %>%
    data.frame() %>%
    mutate_all(funs(case_when(. < X40 ~ X40,
                              . >= X40 ~ .))) %>% as.matrix()
  rownames(est) <- rownames(estorig)
  colnames(est) <- colnames(estorig)
  
  # Expand the err matrix with the self-transition probs
  err <- rbind(1-colSums(est[1:3,]), est[1:3,],
               est[4,], 1-colSums(est[4:6,]), est[5:6,],
               est[7:8,], 1-colSums(est[7:9,]), est[9,],
               est[10:12,], 1-colSums(est[10:12,]))
  rownames(err) <- paste0(rep(c("A","C","G","T"), each=4), "2", c("A","C","G","T"))
  colnames(err) <- colnames(trans)
  # Return
  return(err)
}

# check what this looks like
errF_3 <- learnErrors(
  filtFs,
  multithread = TRUE,
  nbases = 1e8,
  errorEstimationFunction = loessErrfun_mod3,
  verbose = TRUE
)
## 108604125 total bases in 482685 reads from 1 samples will be used for learning the error rates.
## Initializing error rates to maximum possible estimate.
## selfConsist step 1 .
##    selfConsist step 2
##    selfConsist step 3
##    selfConsist step 4
##    selfConsist step 5
##    selfConsist step 6
##    selfConsist step 7
##    selfConsist step 8
##    selfConsist step 9
##    selfConsist step 10


# check what this looks like
errR_3 <- learnErrors(
  filtRs,
  multithread = TRUE,
  nbases = 1e8,
  errorEstimationFunction = loessErrfun_mod3,
  verbose = TRUE
)
## 106190700 total bases in 482685 reads from 1 samples will be used for learning the error rates.
## Initializing error rates to maximum possible estimate.
## selfConsist step 1 .
##    selfConsist step 2
##    selfConsist step 3
##    selfConsist step 4
##    selfConsist step 5
##    selfConsist step 6
##    selfConsist step 7
## Convergence after  7  rounds.
```

Option 4 Alter loess function arguments (weights and span and degree, also enforce monotonicity)
From Jonalim’s comment in benjjneb/dada2#1307
```{r}
library(magrittr)
library(dplyr)

loessErrfun_mod4 <- function(trans) {
  qq <- as.numeric(colnames(trans))
  est <- matrix(0, nrow=0, ncol=length(qq))
  for(nti in c("A","C","G","T")) {
    for(ntj in c("A","C","G","T")) {
      if(nti != ntj) {
        errs <- trans[paste0(nti,"2",ntj),]
        tot <- colSums(trans[paste0(nti,"2",c("A","C","G","T")),])
        rlogp <- log10((errs+1)/tot)  # 1 psuedocount for each err, but if tot=0 will give NA
        rlogp[is.infinite(rlogp)] <- NA
        df <- data.frame(q=qq, errs=errs, tot=tot, rlogp=rlogp)
        
        # original
        # ###! mod.lo <- loess(rlogp ~ q, df, weights=errs) ###!
        # mod.lo <- loess(rlogp ~ q, df, weights=tot) ###!
        # #        mod.lo <- loess(rlogp ~ q, df)
        
        # jonalim's solution
        # https://github.com/benjjneb/dada2/issues/938
        mod.lo <- loess(rlogp ~ q, df, weights = log10(tot),degree = 1, span = 0.95)
        
        pred <- predict(mod.lo, qq)
        maxrli <- max(which(!is.na(pred)))
        minrli <- min(which(!is.na(pred)))
        pred[seq_along(pred)>maxrli] <- pred[[maxrli]]
        pred[seq_along(pred)<minrli] <- pred[[minrli]]
        est <- rbind(est, 10^pred)
      } # if(nti != ntj)
    } # for(ntj in c("A","C","G","T"))
  } # for(nti in c("A","C","G","T"))
  
  # HACKY
  MAX_ERROR_RATE <- 0.25
  MIN_ERROR_RATE <- 1e-7
  est[est>MAX_ERROR_RATE] <- MAX_ERROR_RATE
  est[est<MIN_ERROR_RATE] <- MIN_ERROR_RATE
  
  # enforce monotonicity
  # https://github.com/benjjneb/dada2/issues/791
  estorig <- est
  est <- est %>%
    data.frame() %>%
    mutate_all(funs(case_when(. < X40 ~ X40,
                              . >= X40 ~ .))) %>% as.matrix()
  rownames(est) <- rownames(estorig)
  colnames(est) <- colnames(estorig)
  
  # Expand the err matrix with the self-transition probs
  err <- rbind(1-colSums(est[1:3,]), est[1:3,],
               est[4,], 1-colSums(est[4:6,]), est[5:6,],
               est[7:8,], 1-colSums(est[7:9,]), est[9,],
               est[10:12,], 1-colSums(est[10:12,]))
  rownames(err) <- paste0(rep(c("A","C","G","T"), each=4), "2", c("A","C","G","T"))
  colnames(err) <- colnames(trans)
  # Return
  return(err)
}

# check what this looks like
errF_4 <- learnErrors(
  filtFs,
  multithread = TRUE,
  errorEstimationFunction = loessErrfun_mod4,
  verbose = TRUE
)
## 287310600 total bases in 1276936 reads from 7 samples will be used for learning the error rates.
## Initializing error rates to maximum possible estimate.
## selfConsist step 1 .......
##    selfConsist step 2
##    selfConsist step 3
##    selfConsist step 4
##    selfConsist step 5
##    selfConsist step 6
##    selfConsist step 7
##    selfConsist step 8
##    selfConsist step 9
##    selfConsist step 10
errR_4 <- learnErrors(
  filtRs,
  multithread = TRUE,
  errorEstimationFunction = loessErrfun_mod4,
  verbose = TRUE
)
## 280925920 total bases in 1276936 reads from 7 samples will be used for learning the error rates.
## Initializing error rates to maximum possible estimate.
## selfConsist step 1 .......
##    selfConsist step 2
##    selfConsist step 3
##    selfConsist step 4
##    selfConsist step 5
##    selfConsist step 6
##    selfConsist step 7
##    selfConsist step 8
## Convergence after  8  rounds.
```

Compare error rates 
```{r}
#save.image("~/Desktop/GitProjects/syncoms_experiment/data/dada2/dada2_env_errors.RData")
# Dots in 0 y-axis is an expected behavior, especially with high-quality reads (Q ≥ 30)
plotErrors(errF, nominalQ = TRUE)
plotErrors(errF_1, nominalQ=TRUE)
plotErrors(errF_2, nominalQ=TRUE)
plotErrors(errF_3, nominalQ=TRUE)
plotErrors(errF_4, nominalQ=TRUE) 

plotErrors(errR, nominalQ=TRUE)
plotErrors(errR_1, nominalQ=TRUE)
plotErrors(errR_2, nominalQ=TRUE)
plotErrors(errR_3, nominalQ=TRUE)
plotErrors(errR_4, nominalQ=TRUE)

```

Option 4 is the best, get rid of the other error rates
```{r}
rm(errF, errF_1, errF_2, errF_3, errR, errR_1, errR_2, errR_3)
```

Step 7: Sample Inference
```{r}
dadaFs <- dada(filtFs, err=errF_4, multithread=TRUE)
dadaFs[[1]]

dadaRs <- dada(filtRs, err=errR_4, multithread=TRUE)
dadaRs[[1]]
```

Step 8: Merge Paired Reads 
```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# Inspect the merger data.frame from the first sample
head(mergers[[1]])
save.image("~/dada2_env_mergers.RData")
```

Step 9: Construct sequence table
```{r}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
save.image("~/dada2_env_seq_table.RData")
```

Step 10: Remove Chimeras 
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)

sum(seqtab.nochim)/sum(seqtab)
save.image("~/dada2_env_chimeras.RData")
```
Step 11: Select specific sequence size 
https://earthmicrobiome.org/protocols-and-standards/16s/#:~:text=Amplification%20protocol&text=Run%20amplicons%20from%20each%20sample,is%20~300%E2%80%93350%20bp.
Run amplicons from each sample on an agarose gel. Expected band size for 515F–806R is ~300–350 bp.
```{r}
table(nchar(getSequences(seqtab.nochim)))
#Peak around 255 bp
plot(table(nchar(getSequences(seqtab.nochim))))
#range selected 240-275
seqtab.nochim <- seqtab.nochim[,nchar(colnames(seqtab.nochim))%in% 240:275]
sum(seqtab.nochim)/sum(seqtab)

dim(seqtab.nochim)
```

Step 12: Track reads through the pipeline
```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)

# sum all input reads across samples: 
sum(track[, "input"]) # 33,975,459
# sum all filtered: 
sum(track[, "filtered"]) # 31,110,900
sum(track[, "denoisedF"]) # 27,631,216
sum(track[, "denoisedR"]) # 27,631,128
sum(track[, "merged"]) # 23,988,478
sum(track[, "nonchim"]) # 23,229,069
```

Step 13: Assign taxonomy
```{r}
str(seqtab.nochim)
taxa <- assignTaxonomy(seqtab.nochim, "~/Desktop/FDA_Lab/Experiment1/sequencing/silva_nr99_v138.2_toGenus_trainset.fa.gz", multithread=TRUE)

save.image("~/dada2_env_taxa_genus.RData")
```

