---
title: "MSE1_16SProcessing"
author: "Natalie F"
date: "2025-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#0. Initial Library
```{r, include=FALSE}
library(rlang)
#if (!require("BiocManager", quietly = TRUE))
 #   install.packages("BiocManager")

#BiocManager::install("dada2")
library(dada2); packageVersion("dada2") #[1] '1.38.0'
library(Biostrings)
library(ShortRead)
library(tidyverse)
```


#1. Taxonomy inspect
```{r}
load("~/dada2_env_taxa_genus.RData")
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

Export
```{r}
write.csv(seqtab.nochim, file = "~/asv_table_16S.csv")
write.csv(taxa, file = "~/tax_table_16S.csv")
save.image("~/dada2_env_taxa_genus_complete.RData")
```

#create phyloseq object
Library
```{r}
library(phyloseq); packageVersion("phyloseq") # v 1.54.2
library(readr)
```

# need to extract the OTU table from the phyloseq object that has been filtered (phyloseq_clean2) so that the number of taxa between the OTU table and taxa table match :) 
```{r}
#otu_for_spp <- otu_table(phyloseq_clean2)
```


#1. Input the files and format
```{r}
# Read in taxa asv table
asv <- seqtab.nochim
asv <- t(asv) # transpose (rows <-> columns)  
asv
dim(asv); nrow(asv); ncol(asv) #Outcome: [1] n of rows + columns [1] n of rows [1] n of columns

# Read taxonomy table 
tax <- taxa #file of taxonomy
taxa_names(tax) <- paste0("Seq", seq(ntaxa(tax))) #rename sequences by simpler names

# Transform the taxa table to matrix
dim(tax); ncol(taxa); nrow(tax) # sequences are rows, taxonomic ranks are columns (6)

tax_mat <- as.matrix(tax)
dim(tax_mat); ncol(tax_mat); nrow(tax_mat) 

# Check if asv and tax_mat have the same number of taxa
nrow(asv); nrow(tax_mat) # yes! 
```

#2. Format
```{r}
asv <- t(asv) # transpose (rows <-> columns) 
samples.out <- rownames(asv)
subject <- sapply(strsplit(samples.out, "D"), `[`, 1)

## Create SAMPLE DATAFRAME ##
#Preparing samdf = sample dataframe with features of samples, i.e. site, season, depth, etc
max=read.csv("~/MSE1_metadata.csv", header = TRUE )

saveRDS(max, file = "samdf.rds")

samdf_phy <- readRDS("samdf.rds")
nrow(samdf_phy) #check number of rows

#remove sample ID 59 and 114 (they died during the experiment and have no data associated with them) 
samdf_phy2 <- samdf_phy[-c(59, 114), ]

# make sure rownames of metadata match column names of ASV table 
library(data.table)
setorder(samdf_phy2, cols="ID")
samdf_phy2
samdf_final=cbind(subject, samdf_phy2, rownames=1)
str(asv)
```

#3. Create phyloseq class object ##
#with asv table, taxa table and sample metadata
```{r}
physeq <- phyloseq(otu_table(asv, taxa_are_rows = FALSE), # samples are rows
                   sample_data(samdf_phy2), 
                   tax_table(tax))
```

```{r}
save(physeq, file="~/phyloseq_original.Rdata")
sum(sample_sums(physeq)) #23229069 reads across all samples before filtering
ntaxa(physeq) #259076 ASVs before filtering
```

Remove Chroloplast and or mitochondrial taxa
```{r}
phyloseq_clean <- subset_taxa(physeq, Family!= "Mitochondria" | is.na(Family) & Order!="Chloroplast" | is.na(Order))
sum(sample_sums(phyloseq_clean)) #23023036 reads across all samples 
ntaxa(phyloseq_clean) #255370 ASVs
```

#3. Retain just assigned Bacteria Kingdom
```{r}
phyloseq_clean<-subset_taxa(phyloseq_clean, Kingdom == "Bacteria")
sum(sample_sums(phyloseq_clean)) #21703148 
ntaxa(phyloseq_clean) #253015 ASVs
```

#4. Remove ASV with 2 or less reads - we consider a missreading.
```{r}
phyloseq_clean2 <- prune_taxa(taxa_sums(phyloseq_clean)>2, phyloseq_clean)
sum(sample_sums(phyloseq_clean2)) #21528028 reads 
min(sample_sums(phyloseq_clean2)) # sample with the minimum # of reads: 63 reads (this is sample ID 47) 
```

```{r}
#Use prevalence (number of samples where an ASV is present) instead of total counts. Keep ASVs present in ≥7 samples:

ps <- phyloseq_clean2
margin <- ifelse(taxa_are_rows(ps), 1, 2)               # 1 = taxa rows, 2 = taxa cols
prev   <- apply(otu_table(ps) > 0, margin, sum)         # number of samples with >0
phyloseq_clean2 <- prune_taxa(prev > 7, ps)            # drop ASVs seen in less than 7 samples

ntaxa(phyloseq_clean2) #12313 ASVs 
sum(sample_sums(phyloseq_clean2)) # 20,325,685 reads 

save(phyloseq_clean2, file="~/phyloseq_clean.Rdata")

#That removes any ASV that appears in less than 7 samples (regardless of its read count).
```

#5. Plot rarefaction curve 
```{r}
library(vegan) #v 2.7-3
#rare <-rarecurve(otu_table(physeq), step=50, cex=0.5)

tab <- otu_table(phyloseq_clean2)

class(tab) <- "matrix"

#tab <- t(tab)
rare <-rarecurve(tab, step=50, cex=0.5, label = T)
```

#6. rarefy
Total dataset
pull out otu_table and rarefy as per P. Schloss (2023)
```{r}
selected_rare <- phyloseq_clean2

#select the 
sort(sample_sums(phyloseq_clean2)) #50,000
#rarefy at 50,000

#rarefy
#install.packages('devtools')
#devtools::install_github("GuillemSalazar/EcolUtils")
require(EcolUtils)

asv_tab_reads <- data.frame(otu_table(selected_rare)) #extract
asv_tab_rare <- rrarefy.perm(x = asv_tab_reads , sample = 50000, n = 1000, round.out = T) 

rowSums(asv_tab_rare)

#update otu table in phylsoeq object
otu_table(selected_rare) <-  otu_table(asv_tab_rare, taxa_are_rows = F)

sort(sample_sums(selected_rare))

#compare two permuted otu table to ensure it worked.
View(otu_table(phyloseq_clean2)[1:10, 1:10])
View(otu_table(selected_rare)[1:10, 1:10])
 
```


# Export
```{r}
physeq_rare <- selected_rare

save(physeq_rare, file="~/phyloseq_rarefied.Rdata")
```









