---
title: "MSE1_dada2_cutadapt"
output: html_document
date: "2025-11-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Step 1: Initial Library
```{r, include=FALSE}
library(rlang)
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("dada2")
library(dada2); packageVersion("dada2") #[1] '1.38.0'
library(Biostrings)
library(ShortRead)
library(tidyverse)
```

Step 2: Define the following path variable so that it points to the extracted directory on your machine: 
```{r}
path <- "./raw_reads/"
#Check path
list.files(path)
path
```

Step 3: Read the names of fastq files 
```{r}
# Forward and reverse fastq filenames have format: PROJECT_RH_SAMPLEID_16S_R1_001.fastq and PROJECT_RH_SAMPLEID_16S_R2_001.fastq
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# Extract sample names, assuming filenames have format: PROJECT_RH_SAMPLEID_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 3)

# Inspect read quality profiles - to see if they are corrupted
#The sequencing is 2x300 bp
plotQualityProfile(fnFs[1:2]) # not corrupted
plotQualityProfile(fnRs[1:2]) # not corrupted
```
# On terminal - remove the Poly-G tails
--- install
Cutadapt will be needed for the G-tails from NextSeq
But before, install on terminal:
$ conda create -n dada2 python=3.10 #only have to do this once
$ conda activate dada2
$ conda install -c conda-forge -c bioconda cutadapt
$ cutadapt --version
$ conda install -c bioconda dada2

--- check G-tails
conda activate dada2
cd ./raw_reads
$ run the following on bash
```{r}
# config
N=10000
MIN_TAIL=10
mkdir -p polyG_checks

# choose decompressor (macOS often uses gzcat)
CAT=$(command -v gzcat || command -v zcat)

polyG_check() {
  fq="$1"
  "$CAT" "$fq" 2>/dev/null | \
  awk -v N="$N" -v MIN="$MIN_TAIL" -v FILE="$(basename "$fq")" '
    NR%4==2 {
      gsub(/\r/,"");                # strip CR if present
      seq = toupper($0);            # match like your R (uppercase)
      if (match(seq, /G+$/)) len = RLENGTH; else len = 0;
      t[len]++; total++;
      if (total >= N) exit
    }
    END {
      with=0; maxg=0;
      for (k in t) {
        kk = k + 0
        if (kk >= MIN) with += t[k]
        if (kk > maxg) maxg = kk
      }
      pct = (total>0) ? 100*with/total : 0
      printf("file\ttotal_reads\twith_polyG\tpct_polyG\tmax_G_tail\n");
      printf("%s\t%d\t%d\t%.2f\t%d\n", FILE, total, with, pct, maxg);
    }' > "polyG_checks/$(basename "$fq").tsv"
}

export -f polyG_check
export CAT N MIN_TAIL

# run on all fastqs
for f in *_R1_001.fastq.gz *_R2_001.fastq.gz; do
  polyG_check "$f" &
done
wait

# peek
head -n 2 polyG_checks/*.tsv
```
--- remove Poly G
```{r}
CUTADAPT="~/miniconda3/envs/dada2/bin/cutadapt"
mkdir -p cutadapt

for r1 in *_R1_001.fastq.gz; do
  r2="${r1/_R1_/_R2_}"                   # MSE1_RH_1_16S_S147_R2_001.fastq
  sample="$(printf '%s\n' "$r1" | cut -d'_' -f3)"       # sample ID token is after the first two underscores
  out1="cutadapt/${sample}_R1_cut.fastq"
  out2="cutadapt/${sample}_R2_cut.fastq"
  logf="cutadapt/${sample}_polyG.log"

  "$CUTADAPT" --nextseq-trim=20 --max-n 0 --minimum-length 35 \
    -o "$out1" -p "$out2" \
    "$r1" "$r2" > "$logf" 2>&1

  echo "Done: $sample"
done
```


Step 4: in R: Check that poly-G tails were removed by cutadapt (in R)
```{r}
path_g <- file.path(path, "cutadapt")   #  where the *_cut.fastq live

fnFs.g <- sort(list.files(path_g, pattern = "_R1_cut\\.fastq$", full.names = TRUE))
fnRs.g <- sort(list.files(path_g, pattern = "_R2_cut\\.fastq$", full.names = TRUE))

length(fnFs.g); length(fnRs.g)          # must be equal and > 0

# Inspect read quality profiles - to see if they are corrupted
#The sequencing is 2x300 bp
plotQualityProfile(fnFs.g[2]) # not corrupted
plotQualityProfile(fnRs.g[2]) # not corrupted
```

# On terminal - Check if the Poly-G tails were removed

Check whether it is completed and correct
```{r}
cd cutadapt
N=10000
MIN_TAIL=10
mkdir -p polyG_checks

# choose decompressor (macOS often uses gzcat)
CAT=$(command -v gzcat || command -v zcat)

polyG_check() {
  fq="$1"
  "$CAT" "$fq" 2>/dev/null | \
  awk -v N="$N" -v MIN="$MIN_TAIL" -v FILE="$(basename "$fq")" '
    NR%4==2 {
      gsub(/\r/,"");                # strip CR if present
      seq = toupper($0);            # match like your R (uppercase)
      if (match(seq, /G+$/)) len = RLENGTH; else len = 0;
      t[len]++; total++;
      if (total >= N) exit
    }
    END {
      with=0; maxg=0;
      for (k in t) {
        kk = k + 0
        if (kk >= MIN) with += t[k]
        if (kk > maxg) maxg = kk
      }
      pct = (total>0) ? 100*with/total : 0
      printf("file\ttotal_reads\twith_polyG\tpct_polyG\tmax_G_tail\n");
      printf("%s\t%d\t%d\t%.2f\t%d\n", FILE, total, with, pct, maxg);
    }' > "polyG_checks/$(basename "$fq").tsv"
}

export -f polyG_check
export CAT N MIN_TAIL

# run on all fastqs
for f in *_R1_cut.fastq *_R2_cut.fastq; do
  polyG_check "$f" &
done
wait

# peek
head -n 2 polyG_checks/*.tsv # should all be 0
```

# In R - check primers
Primer sequence set up  
```{r}
FWD <- "GTGYCAGCMGCCGCGGTAA"  ## this is 515f #length 19
REV <- "GGACTACNVGGGTWTCTAAT"  ## this is 806Br #length 20

#Check the primers
allOrients <- function(primer) {
    # Create all orientations of the input sequence
    require(Biostrings)
    dna <- DNAString(primer)  # The Biostrings works w/ DNAString objects rather than character vectors
    orients <- c(Forward = dna, Complement = Biostrings::complement(dna), Reverse = Biostrings::reverse(dna),
        RevComp = Biostrings::reverseComplement(dna))
    return(sapply(orients, toString))  # Convert back to character vector
}

FWD.orients <- allOrients(FWD)
REV.orients <- allOrients(REV)
FWD.orients
```


Step 5: Remove primers:
```{r}
FWD <- "GTGYCAGCMGCCGCGGTAA"  ## this is 515f #length 19
REV <- "GGACTACNVGGGTWTCTAAT"  ## this is 806Br #length 20

#Check the primers
allOrients <- function(primer) {
    # Create all orientations of the input sequence
    require(Biostrings)
    dna <- DNAString(primer)  # The Biostrings works w/ DNAString objects rather than character vectors
    orients <- c(Forward = dna, Complement = Biostrings::complement(dna), Reverse = Biostrings::reverse(dna),
        RevComp = Biostrings::reverseComplement(dna))
    return(sapply(orients, toString))  # Convert back to character vector
}

FWD.orients <- allOrients(FWD)
REV.orients <- allOrients(REV)
FWD.orients
```
We are now ready to count the number of times the primers appear in the forward and reverse read, while considering all possible primer orientations. Identifying and counting the primers on one set of paired end FASTQ files is sufficient, assuming all the files were created using the same library preparation, so we’ll just process the first sample.

```{r}
primerHits <- function(primer, fn) {
    # Counts number of reads in which the primer is found
    nhits <- vcountPattern(primer, sread(readFastq(fn)), fixed = FALSE)
    return(sum(nhits > 0))
}

rbind(FWD.ForwardReads = sapply(FWD.orients, primerHits, fn = fnFs.g[[2]]), FWD.ReverseReads = sapply(FWD.orients,
    primerHits, fn = fnRs.g[[2]]), REV.ForwardReads = sapply(REV.orients, primerHits,
    fn = fnFs.g[[2]]), REV.ReverseReads = sapply(REV.orients, primerHits, fn = fnRs.g[[2]]))
```
# On terminal - Remove primers
```{r}
# cd to your FASTQ directory
cd ./raw_reads/

mkdir -p cutadapt_primers

FWD="GTGYCAGCMGCCGCGGTAA"      # 515F
REV="GGACTACNVGGGTWTCTAAT"     # 806R
FWD_RC="TTACCGCGGCKGCTGRCAC"   # RC(515F)
REV_RC="ATTAGAWACCCBNGTAGTCC"  # RC(806R)

for r1 in cutadapt/*_R1_cut.fastq; do
base=$(basename "$r1")  
s=${base%_R1_cut.fastq}
  r2="cutadapt/${s}_R2_cut.fastq"
  out1="cutadapt_primers/${s}_R1_cut2.fastq"
  out2="cutadapt_primers/${s}_R2_cut2.fastq"
  logf="cutadapt_primers/${s}_primer.log"

  cutadapt \
    -g "${FWD}"  -a "${REV_RC}" \
    -G "${REV}"  -A "${FWD_RC}" \
    -n 2 \
    --overlap 17 \
    --pair-filter=both \
    --minimum-length 100 \
    -o "$out1" -p "$out2" \
    "$r1" "$r2" > "$logf" 2>&1

  echo "Primer-trimmed: $s"
done
```

Step 6:  in R: Check that primers were removed: 
```{r}
path_g.p <- file.path(path, "cutadapt_primers")   # <- where the *_cut.fastq live
list.files(path_g.p)

fnFs.g.p <- sort(list.files(path_g.p, pattern = "_R1_cut2\\.fastq$", full.names = TRUE))
fnRs.g.p <- sort(list.files(path_g.p, pattern = "_R2_cut2\\.fastq$", full.names = TRUE))

# Inspect read quality profiles - to see if they are corrupted
#The sequencing is 2x300 bp
plotQualityProfile(fnFs.g.p[1:2]) # not corrupted
plotQualityProfile(fnRs.g.p[1:2]) # not corrupted

rbind(FWD.ForwardReads = sapply(FWD.orients, primerHits, fn = fnFs.g.p[[1]]), FWD.ReverseReads = sapply(FWD.orients,
    primerHits, fn = fnRs.g.p[[1]]), REV.ForwardReads = sapply(REV.orients, primerHits,
    fn = fnFs.g.p[[1]]), REV.ReverseReads = sapply(REV.orients, primerHits, fn = fnRs.g.p[[1]]))
```
